---
title: 'Tidy Evaluation'
date: '2026-02-26'
engine: knitr
type: slides
execute:
  error: true
  eval: true
categories:
- Week06
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(rlang)
```

# Tidy Evaluation is comprised of

- quasiquotation,

- quosures,

- and data masks


# Quosures

Idea: capture an expression with its environment (for later evaluation)

- functions `enquo` and `enquos` (and lesser used, `quo` and `quos`): 

```{r}
quo(x <- 2*exp(1))

f <- function(x) enquo(x)

f(a+b+d/2)
```



- also: `new_quosure`

```{r}
qu <- new_quosure(expr(x^2+y), env(x = 2, y = 1))
qu
eval_tidy(qu)
```
# Quosures 

```{r}
f <- function(...) {
  x <- 1
  g(..., f = x)
}
g <- function(...) {
  enquos(...)
}

x <- 0
qs <- f(global = x)
qs
```
#

```{r}
q4 <- new_quosure(expr(x + y + z))

q4
```

```{r}
q4 |> get_expr()
q4 |> get_expr() |> expr_text()

q4 |> get_env()
```

# Your Turn  {background-color="#006666"}
[From Advanced R](https://adv-r.hadley.nz/evaluation.html#exercises-62):

1. Predict what each of the following quosures will return if evaluated.

```{r eval = FALSE}
x <- 0
q1 <- new_quosure(expr(x), env(x = 1))
q2 <- new_quosure(expr(x + !!q1), env(x = 10))
q3 <- new_quosure(x + eval_tidy(q2), env(x = 100))
```

2. Write an `enenv()` function that captures the environment associated with an argument. (Hint: this should only require two function calls.)


# 

```{r}
enenv <- function(x) {
  qu <- enquo(x)
  get_env(qu)
}

enenv(q3)
```
# Data Masks

Sometimes we are dealing with expressions that contain both variables (symbols from a data frame) and objects from an environment

In evaluations we need to consider both the data frame and the environment

The function eval_tidy has `data` as a third argument; if it is specified, it will be used first when evaluating expressions:

```{r}
eval_tidy(expr(mpg), data = mtcars)
```

# Tidyverse programming

Data masking introduces some ambiguity about the environment in which each object lives, e.g

```
1/mpg / mile_to_100km * gallons_to_liter
```

In the `.data` or in the parent environment `.env`?

```{r}
gallons_to_liter =  3.79
mile_to_100km = 1.60934 /100

mtcars |> 
  dplyr::mutate(
    lper100km = 1/mpg / mile_to_100km * gallons_to_liter
) |> dplyr::select(mpg, lper100km) |> dplyr::arrange(lper100km)
```

# Data Pronouns

`.data$x` is always the x variable in the data frame `.data`

`.env$x` is always the x object in the environment `.env`

When data masking is used in programming, the expectation of R CMD check is that the ambiguity is resolved

e.g.:

```
  ghc: no visible binding for global variable ‘cluster’
```
is resolved by using `.data$cluster` in the code (this introduces a dependence to the `rlang` package)


# Your Turn  {background-color="#006666"}

Can we implement a function `my_select` now that works the same way as `dplyr::select`? 

```{r eval=FALSE}
my_select <- function (.data, ...) {

}
```

# 

From [Advanced R](https://adv-r.hadley.nz/evaluation.html#select)

```{r}
select2 <- function(data, ...) {
  dots <- enquos(...)

  vars <- as.list(set_names(seq_along(data), names(data)))
  cols <- unlist(purrr::map(dots, eval_tidy, vars))

  data[, cols, drop = FALSE]
}
```

```
select2(mtcars, 1:2, "wt", dplyr::starts_with("d"))
```

Still a no.

# `my_select` with a cheat

```{r}
my_select <- function (.data, ...) {
  vars <- enquos(...)
  
  res <- lapply(vars, FUN = function(x) {
    e <- get_expr(x)
    if (is_symbol(e)) e <- as_string(e)
    if (is_call(e)) e <- tidyselect::eval_select(e, data=.data)
    `[`(.data, e) 
  }) |> data.frame()
}
```

```{r}
mtcars |> dplyr::select("mpg", cyl, 3, tidyr::starts_with("d")) |> head()
```
