---
title: 'Unit Testing'
date: '2026-03-05'
engine: knitr
type: slides
execute:
  error: true
  eval: false
categories:
- Week07
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---


# Outline

- Motivation

- Automated Testing and Unit Tests

- Testing Functions in R with `testthat`

- Coverage

- Package Checks 

# Informal Testing


- testing is part of function writing

- sometimes we forget to test a special case

- also: functions can create side effects and introduce new errors


# Automate Testing with Unit Tests

- Unit tests are tests that evaluate one feature of a function



- run automatically after you make changes to the code



- Unit tests are one part of the larger framework of _Test Driven Development_

# Test Driven Development

_Test Driven Development_ is an approach to programming where the unit tests are written before the actual code:

- Requirements are decided in advance

- Code must meet the requirements (and only the requirements)

# Why use unit tests? {.smaller}

For a bit of extra work, you get: 

- Fewer bugs
    - Essential functions are tested and issues are caught quickly
    - Visual confirmation that essential features are working
    - When debugging, write a test to pinpoint the bug (and stop it from reoccurring)

- Better code structure
    - Modular code: do only one thing per function
    - Document functions as you write them to satisfy tests

- Robust code 
    - make big changes quickly and without downstream problems
    - Any changes that break things should be caught with tests

- (in theory) Easier to start after a break:    
pick up at the last failed test and write code to satisfy that test


# Tools: `testthat`

`testthat`: unit testing for R packages
  
  - structured R package testing
  
  - provides functions to set up and tear down testing environments
  
  - runs each set of tests in a clean environment
  
  - reports whether each test passes or fails

# Tools: `usethis`

`usethis`: helper functions for R package development

  - set up `testthat` for your package: `usethis::use_testthat()`
  
  - create new test files: `usethis::use_test("test1")` creates a new test file

<br>
Note: `testthat` works with packages. [This github issue](https://github.com/r-lib/testthat/issues/659) has a good discussion of testing outside of the package framework


# Package Testing Workflow

1. Modify your code or tests

2. Test your package (`Ctrl/CMD - Shift - T`)

3. Repeat until all tests pass


# Your Turn {background-color="#006666"}

```{r, eval = 2}
install.packages(c("testthat", "usethis"))
library(testthat)
```

1. Install the `testthat` package for unit testing and the `usethis` package (helper functions to make package development easy)

2. Set the `happyR` package up to work with `testthat`

```{r, eval = F}
# Run create_package to add package infrastructure 
# Adjust the path to refer to your current project directory
usethis::use_testthat()
```

What does your file structure look like now? 



# Testing Structure

A test file consists of 

1. context - a description of the test blocks in the file

2. one or more test blocks:    
`test_that(description, {test statements})`

```{r, eval = F}
context("test-basics")

# Test block
test_that(
  # description
  "multiplication works",
  { # test statements inside this block
    expect_equal(2 * 2, 4)
  }
)
```

# Test Statements {.smaller}

`testthat` has a series of expectation functions:

function | description
---- | ----
`expect_equal(obj, value)` | Is the object equal to a value?
`expect_error(expr)` | Does the expression produce an error?
`expect_gt(obj, value)` | Is the object greater than the value?
`expect_length(obj, value)` | Does the object have length value?

See more with `help(package = "testthat")`

These functions are silent if the expectation is met, and throw an error otherwise.

Expectations are used to construct tests. 


# Your Turn - Writing Expectations {background-color="#006666"}

1. Create a test file with `usethis::use_test("hello")`

2. Write a set of expectations that test the `hello` function


# Test-Driven Development Workflow

1. Write out the specifications for your function

2. Create your test file `usethis::use_test("foo")`

3. Write tests for your function in `tests/testthat/test_foo.R`

4. Write your function and documentation in `R/foo.R`

5. Test your package: `Ctrl/CMD - Shift - T` or use the Build Menu


# Test-Driven Development Example

Function: `mymean(x, na.rm)`

Required Outcome: 

  - calculate average for numeric vector x
  - if `na.rm` is true, missing values in x should be removed before calculating the average

# Test-Driven Development

Function: `mymean(x, na.rm)`

Specific Behavior:

- issue a warning when x is not numeric and return `NA`
- return `NA` when x has `NA`s and `na.rm = F`
- return a value equal to `sum(x)/length(x)` when x has no `NA` values 
- return a value equal to `sum(x2)/length(x2)` where    
`x2 = x[!is.na(x)]` if `na.rm = T`


# Test-Driven Development Example

Important behaviors: 


- issue a warning when x is not numeric and return NA

```{r, eval = F}
x <- letters[1:3]
expect_warning(mymean(x)) # This is the minimal test
expect_warning(mymean(x), # This tests for a specific warning
               "argument is not numeric or logical: returning NA") 

# This tests the value and that a warning is generated
expect_warning(
  expect_true(is.na(mymean(x))) # Function returns NA
)

```


# Test-Driven Development Example

- return `NA` when x has `NA`s and `na.rm = F`
```{r, eval = F}
# Can also test to see if a test fails...
expect_failure(
  expect_equal(mymean(c(1:8, NA), na.rm = F), mymean(c(1:8)))
)

# Or test if something returns NA
expect_true(is.na(mymean(c(1:8, NA))))
```


# Test-Driven Development Example

- return a single numeric value equal to `sum(x)/length(x)`

```{r, eval = F}
x <- 1:8
y <- c(1:8, NA)

# Test that mean(1:8) returns a numerically correct response
expect_equal(mymean(x), 4.5)

# Test that mean(c(1:8, NA)) equals mean(1:8) when na.rm = T
expect_equal(mymean(y, na.rm = T), mymean(x))
```


# Unit tests are...

- Modular (by design)

- Quick to run

- Run in an clean environment<br/>    
    - Set up and Tear down tasks set up the environment for testing
    
- Independent - don't require outside files (traditionally)


# Unit tests

> But, we're statisticians. What about the data?

- use tiny test data (`dput()` is helpful)

- read in toy data from files in the test directory

- use data included with the package

- use data included in base R

- download data from elsewhere with a set-up function, delete it with a tear-down function
    - tests will fail without internet or if the site is down

# Code Coverage: `covr` {.smaller}

> How many unit tests are enough? Is everything tested?

- `covr` is a package that will:
    1. build your package in a clean environment 
    2. run your tests
    3. determine how many times each line was evaluated (through [magic](https://cran.r-project.org/web/packages/covr/vignettes/how_it_works.html))
    4. launch a Shiny app to show you line-by-line coverage reports
    
- 100% coverage is good, but unit tests aren't everything
    - Some lines aren't worth testing
    - Integration testing matters too!

```{r, eval = F}
covr::report() # Run a local code coverage report

```

# Your Turn   {background-color="#006666"}

1. Run a code coverage report locally for your `happyR` package using `covr::report()`

2. Run `use_tidy_github_actions()`. 
This will activate all three main workflows: checking the package, running a code coverage report and re-publishing the website

