---
title: 'Functions in C++'
date: '2026-03-30'
engine: knitr
type: slides
execute:
  error: true
categories:
- Week09
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---


# Connecting R and C++

- lots of support for C++ programming from R community

- on your machine: C++ compiler (check with `which cpp` in the Terminal)

# A first program

```{r}
library(Rcpp)

cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')


add(1,2,3)
```
# Vector input, scalar output

```{r}
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}
```

# Vector input, scalar output in C++

```{r}
cppFunction('double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}')
```


# Package `bench`

```{r}
x <- runif(1e3)
bench::mark(
  sum(x),
  sumC(x),
  sumR(x)
)[1:6]
```
# Let's try this {background-color="#006666"}

Buffon's needle is a Monte Carlo estimation of $pi$ by counting the number of times that a needle of length $\ell$ tossed on a piece of lined paper crosses a line, when the lines are $\ell$ apart. 
The theoretical probability that any tossed needle crosses a line is $2/\pi$.

Write a function `rbuffon(n)` that produces a binary vector with $n$ 1s and 0s (1s for crossing needles and 0s otherwise).

How many digits of $\pi$ can be reliably estimated with 50,000 tosses?


Double check the digits of $\pi$ against [Eve Andersson's one million digits of pi](https://github.com/eneko/Pi/blob/master/one-million.txt)


# Let's try this {background-color="#006666"}

Use [Advanced R, Chapter 25](https://adv-r.hadley.nz/rcpp.html) to implement a C version of `rbuffon`. 

How do the functions compare in speed?

What about number of crossings in n throws?

# Buffon's Needle in R and C

```{r}
rbuffon <- function(n) {
  # Simulate the center of the needle's distance from the nearest line
  # Uniform on [0, 0.5] since distance between lines = 1
  center_dist <- runif(n, min = 0, max = 0.5)
  
  # Simulate the acute angle the needle makes with the lines
  # Uniform on [0, pi/2]
  angle <- runif(n, min = 0, max = pi / 2)
  
  # A needle crosses a line if the horizontal projection of its half-length
  # is >= the distance from its center to the nearest line.
  # Half-length projection = 0.5 * sin(angle), needle length = 1
  crosses <- as.integer(0.5 * sin(angle) >= center_dist)
  
  return(crosses)
}

2/mean(rbuffon(50000))
2/mean(rbuffon(50000))
2/mean(rbuffon(50000))
```
# C implementation

download [code file](../code/buffons-needle.cpp)

```{r}
library(Rcpp)

# Compile and load the C++ function
sourceCpp("../code/buffons-needle.cpp")

2/mean(rbuffon_cpp(50000))
2/mean(rbuffon_cpp(50000))
2/mean(rbuffon_cpp(50000))

```
#

```{r}
bench::mark(
  2/mean(rbuffon(100000))
)[1:6]

bench::mark(
  2/mean(rbuffon_cpp(100000))
)[1:6]

```
#

```{r}
bench::mark(
  sum(rbuffon(1000000))
)[1:6]

bench::mark(
  rbuffon_count_cpp(1000000)
)[1:6]

```

# Example: Spigot algorithm for $\pi$

```{r}
# install.packages("Rcpp")
library(Rcpp)

cppFunction('
#include <Rcpp.h>
#include <boost/multiprecision/cpp_int.hpp>
using namespace Rcpp;
using boost::multiprecision::cpp_int;

// [[Rcpp::export]]
std::string pi_spigot_boost(int digits) {
    cpp_int q = 1, r = 0, t = 1;
    cpp_int k = 1, n = 3, l = 3;

    int count = 0;
    std::string out = "";
    int newline = 0;

    while (count < digits) {
        if ((4*q + r - t) < n * t) {
            // emit digit
            out += n.convert_to<std::string>();
            count++;
            newline++;

            if (count == 1) {
              out += ".\\n";
              newline = 0;
            }
            if (newline==50) {
              out += "\\n";
              newline = 0;
            }

            cpp_int q_old = q;
            cpp_int r_old = r;

            q = 10 * q_old;
            r = 10 * (r_old - n * t);
            n = (10 * (3*q_old + r_old) / t) - 10 * n;
        } else {
            cpp_int q_old = q;
            cpp_int r_old = r;
            cpp_int t_old = t;
            cpp_int k_old = k;
            cpp_int l_old = l;

            q = q_old * k_old;
            r = (2*q_old + r_old) * l_old;
            t = t_old * l_old;
            k = k_old + 1;
            n = (q_old * (7*k_old) + 2 + r_old * l_old) / (t_old * l_old);
            l = l_old + 2;
        }
    }

    return out;
}
')

# test
cat(pi_spigot_boost(101))

```
