---
title: 'Object Oriented Programming (OOP) in R'
date: '2026-04-02'
engine: knitr
type: slides
execute:
  error: true
categories:
- Week09
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---

```{r echo=FALSE, include=FALSE}
library(tidyverse)
```

# Outline

- What is Object Oriented Programming? 

- OOP models in R

- Constructors 

# Object Oriented Programming

Idea of object oriented programming is that the user can define abstract objects (**classes**, e.g. a linear model object) that contains named 'pieces' (**fields**, e.g. the data, the parameter estimates, residuals, ...) and associated 'methods' (functions, e.g. `predict`)

**Polymorphism** allows us to 'over-write' existing methods and implement specialized functionality that way

Methods have the form `class.methodname` - at the core of R and lazy evaluation: e.g. `summary`, `print`, ...

# OOP models in R

In chronological order:

- base 

- S3 

- S4 

- R6

Check any object in R for its class/type with `typeof` or `sloop::otype`


# OOP models used in R


```{r}
typeof(1:10)
sloop::otype(1:10)
```
`ggplot2` is a based on `grid` and uses `S3`:

```{r}
gg <- diamonds |> ggplot(aes(x = carat, y = price)) + geom_point()
sloop::otype(gg)
```

`shiny` uses `R6`

`lme4`, `stats4` use `S4`

