---
title: 'The S3 Object model'
date: '2026-04-07'
engine: knitr
type: slides
execute:
  error: true
  cache: false
categories:
- Week09
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---

```{r echo=FALSE, include=FALSE}
library(tidyverse)
```

# Outline

- Constructors and Validators

- Inheritance

- Methods 

# S3 

make an `S3` object: give an object a class:

```{r}
data <- matrix()
class(data) <- "simMatrix"

sloop::otype(data)
```
```{r}
data <- structure(matrix(), class = "simMatrix")

sloop::otype(data)
```
Obviously this system is very exploitable -- and accident-prone

# Working with S3 relies on good practice

Just claiming some object is of a particular class doesn't make it so, and it could lead to terrible problems.

```{r  warning=TRUE}
x <- as.Date("2026-04-07")
str(x)

y <- "2026-04-07"
class(y) <- "Date"
str(y)
```

# Constructors 


Good practice, but not required: *constructor* `new_classname`

```{r cache=FALSE}
new_simMatrix <- function(size = NULL) {
  if (is.null(size)) {
      size <- 0
  }
  dimnames= list(vector("character", size), vector("character", size))
  if (size > 0) dimnames <- list(as.character(1:size), as.character(1:size))
  structure(matrix(numeric(length= size*size), nrow=size, ncol=size,
                   dimnames = dimnames), 
            class = "simMatrix")
}

new_simMatrix(size = 2)
```



# Validator

*Validator*: check that object is 'right'

```{r}
validate_simMatrix <- function(x) {
  # simMatrix is supposed to be a square matrix
  dims <- dim(x) 
  stopifnot(dims[1]==dims[2]) # matrix is supposed to be square
  values <- unclass(x)
  stopifnot(is.numeric(values)) 
  x
}
```


# Helper for users

Focus here on the user, i.e. think about defaults and use cases


**Constructor** of the form `class()` to create a new object of class `class`


Providing support for **generics**, such as `print`

# Constructor

Idea: create a new `simMatrix` and validate it

```{r}
simMatrix <- function(size = NULL) {
  x <- new_simMatrix(size)
  validate_simMatrix(x)  
}
```

`new_simMatrix` only constructs one object at a time - can't use that for casting a vector

# Vectorized Constructor

```{r}
simMatrix_one <- function(size = NULL) { # rename previous constructor
  x <- new_simMatrix(size)
  validate_simMatrix(x)  
}

simMatrix <- function(x) { # x can now be a vector of objects 
 lapply(x, FUN = simMatrix_one)
}
```

# Example

constructor looks like a cast

```{r}
mtcars <- mtcars |> mutate(
  carb = simMatrix(carb)
)

head(mtcars)
```

# 

look closer, we see that there is now a list vector with `simMatrix` objects

```{r}
str(head(mtcars))
```



# Which generic is used?

```{r}
x <- new_simMatrix(size = 3)

sloop::s3_dispatch(print(x))
```



# Inheritance

in 90% of situations classes aren't started from scratch but use inheritances:

```{r}
x <- new_simMatrix(1)
class(x) <- c("simMatrix", "matrix") # order matters: more specific first
```

creates a new object of `new class` that inherits all of the properties from the parent class `simMatrix`

that's how `ggplot2` extensions work (new `geom` and `stat`)

```{r}
sloop::s3_dispatch(print(x))
```

# Providing generic support


```{r}
print.simMatrix <- function(x) {
  dims <- dim(x)
  cat("simMatrix <", dims[1]," x ",dims[2], ">\n")
  print(unclass(x))
}

sloop::s3_dispatch(print(x))
```
```{r}
head(mtcars)
```

# 

```{r}
print(mtcars$carb[1])

print(head(mtcars$carb))
```

# Your Turn  {background-color="#006666"}

The treatment of colors in R is ... strange. 
The `colors()` object is a vector of `r length(colors())` named colors 

```{r}
head(colors())
```
These named colors, together with hex colors from RGB provide the colors used throughout
R's plotting routines.

Create an S3 class "color" that saves a color's RGB hexcode (`col2rgb`) together with its name, if given.

# Your Turn {background-color="#006666"}

Expand the color class by adding a second color for annotations - either in white or in black, depending on which color's contrast is bigger `colorspace::contrast_ratio`

# A different constructor

By default `simMatrix` is called with a size argument. 

What, if the input has the form of a `data.frame` with columns for `row.names`, `column.names` and `value`?

```{r}
dframe <- data.frame(expand.grid(rows=paste0("L",1:6), cols=paste0("L",1:6)), 
                     value = runif(36))

simMatrix(dframe)
```
# Writing methods

```{r}
simMatrix <- function(x, ...) {
  UseMethod("simMatrix")
}
simMatrix.numeric <- function(x, ...) {
  # interpret integer as size object
  new_simMatrix(size = x)
}

simMatrix.data.frame <- function(x, formula = value ~ rows+cols, ...) {
  mat <- xtabs(formula, data = dframe)
  class(mat) <- "simMatrix"
  validate_simMatrix(mat)
}
```

# Now use the different constructors

```{r}
simMatrix(2L)
simMatrix(dframe)
```
