---
title: 'The S3 Object model'
date: '2026-04-07'
engine: knitr
type: slides
execute:
  error: true
  cache: false
categories:
- Week09
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---

```{r echo=FALSE, include=FALSE}
library(tidyverse)
```

# An example 

```{r}
colorspace::contrast_ratio(palette(), "white", plot = TRUE)
```

# 

```{r}
colorspace::contrast_ratio(palette(), "black", plot = TRUE)
```

# Your Turn  {background-color="#006666"}

The `colorspace` package implements color objects using the S4 model in R.

Find out how that looks like from the source code.

# S4 addresses limitations of S3

Why S4?

  - Formal class definitions
  - Stronger type safety
  - Multiple dispatch


# Defining S4 Classes

```r
setClass(
  "Person",
  slots = list(
    name = "character",
    age  = "numeric"
  )
)
```

Slots define structure and types

# Common Constructor

```r
p <- new("Person", name = "Alice", age = 30)
```

Accessing Slots with `@`

```r
p@name
p@age
```

# Validity and Type Checking

- Automatic type enforcement

```r
setValidity("Person", function(object) {
  if (object@age < 0) "Age must be non-negative"
  else TRUE
})
```

- Ensures consistent objects

## Generics

```r
setGeneric("greet", function(object)
  standardGeneric("greet")
)
```

* Defines interface

---

## Methods

```r
setMethod("greet", "Person", function(object) {
  paste("Hello", object@name)
})
```

```r
greet(p)
```

---

## Multiple Dispatch

```r
setGeneric("combine", function(x, y)
  standardGeneric("combine")
)

setMethod("combine", c("Person", "Person"),
  function(x, y) {
    paste(x@name, "and", y@name)
  }
)
```

* Dispatch based on multiple arguments

---

## Inheritance

```r
setClass("Employee",
  slots = list(id = "numeric"),
  contains = "Person"
)
```

* Inherits slots and methods

---

## Accessor Functions (Best Practice)

```r
setGeneric("name", function(object)
  standardGeneric("name")
)

setMethod("name", "Person", function(object)
  object@name
)
```

* Avoid direct slot access

---

## Interoperability

* Use S3 with S4:

```r
setOldClass("data.frame")
```

* Mixed systems are common

---

## Advantages and Tradeoffs

### Advantages

* Formal structure
* Strong validation
* Multiple dispatch

### Tradeoffs

* Verbose
* Steeper learning curve

---

## Summary

* S4 = formal OOP in R
* Key features:

  * Typed slots
  * Explicit classes
  * Multiple dispatch
* Best for complex systems

---

## Exercises

1. Define class `Rectangle` with slots `width`, `height`
2. Add validity check (positive values)
3. Write method `area()`
4. Extend to `Square`
5. Implement `combine()` for rectangles
