---
title: 'The S4 Object model'
date: '2026-04-09'
engine: knitr
type: slides
execute:
  error: true
  cache: false
categories:
- Week09
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---

```{r echo=FALSE, include=FALSE}
library(tidyverse)
```

# An example 

```{r}
colorspace::contrast_ratio(palette(), "white", plot = TRUE)
```

# 

```{r}
colorspace::contrast_ratio(palette(), "black", plot = TRUE)
```

# Your Turn  {background-color="#006666"}

The `colorspace` package implements color objects using the S4 model in R.

Find out how that looks like from the source code.

# S4 addresses limitations of S3

Why S4?

  - Formal class definitions
  - Stronger type safety
  - Multiple dispatch


# Defining S4 Classes

```{r}
Person <- setClass(
  "Person",
  slots = list(
    name = "character",
    age  = "numeric"
  )
)
```

Slots define structure and types

# Common Constructor

use `new` or the name of the class:

```{r}
alice <- new("Person", name = "Alice", age = 27)
bob <- Person(name="Bob", age = 25)
```

Accessing Slots with `@`

```{r}
alice@name
bob@age
```

# Validity and Type Checking

- Automatic type enforcement

```{r error=TRUE}
setValidity("Person", function(object) {
  if (object@age < 0) "Age must be non-negative"
  else TRUE
})

cynthia <- Person(name = "Cynthia", age = -2)

```

- Ensures consistent objects

# Generic Functions

```{r}
setGeneric("greet", function(object)
  standardGeneric("greet")
)
```

* Defines interface

# Methods

```{r}
setMethod("greet", "Person", function(object) {
  paste("Hello, ", object@name)
})
```

```{r}
greet(alice)
```
# Multiple Dispatch

```{r}
setGeneric("combine", function(x, y)
  standardGeneric("combine")
)

setMethod("combine", c("Person", "Person"),
  function(x, y) {
    paste(x@name, "and", y@name)
  }
)

combine(alice, bob)
```

* Dispatch based on multiple arguments



# Inheritance

```{r}
setClass("Employee",
  slots = list(id = "numeric"),
  contains = "Person"
)
```

* Inherits slots and methods

# Accessing Parts

```{r}
alice |> str()

alice@age
```

better: set up accessor functions:

```{r}
setGeneric("name", function(object)
  standardGeneric("name")
)

setMethod("name", "Person", function(object)
  object@name
)

name(bob)
age(alice)
```

* Avoid direct slot access

# Interoperability

* Use S3 with S4:

```{r}
setOldClass("data.frame")
```

* Mixed systems are common

# Summary

* S4 is a more formal OOP model than S3
* Key features:

  * Typed slots
  * Explicit classes
  * Multiple dispatch
* Better for complex problems

At the expense of steeper learning curve, more verbose function

# Your Turn  {background-color="#006666"}

1. Define class `Rectangle` with slots `width`, `height`
2. Add validity check (positive values)
3. Write method `area()`
4. Extend to `Square`
5. Implement `combine()` for rectangles


# Your Turn  {background-color="#006666"}

1. Create the class `RectangleS3` using S3 with constructor `height` and `widths`. 
2. Add validity check (positive values)
3. Write method `area()`
4. How does the method `combine` have to be adjusted?

5. Create 5,000 rectangles from a normally distributed heights and widths, check that they are valid and calculate their area. 
6. Compare the speed of the two implementations. 

