---
title: "Parallel Computing in R"
date: '2026-04-21'
engine: knitr
type: slides
execute:
  error: true
  cache: false
categories:
- Week12
format:
  revealjs:
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---

```{r setup, echo=FALSE}
library(tidyverse)
```

# Idea of Parallelization

- by default R is using only a single processor (core)

- most modern machines have multiple processors

- parallel code makes use of these multiple processors

- challenge: identify those pieces of code that can be run in parallel


# Types of Parallelism

__Data Parallelism__

- Same operation on different data chunks
- Example: bootstrap

__Task Parallelism__

- Different tasks in parallel
- Example: multiple models

# Embarassingly parallel processes

An *embarassingly parallel process* is defined as a process where 
little or no effort is needed to separate the problem into a number of parallel tasks

Loops and grouping structures are prime candidates:

```{r eval = FALSE}
dataframe %>% group_by(group) %>%
  mutate(<<some calculation>>)
```

we can exploit group structure to identify code pieces that can be parallelized


# Parallel Architecture

__Shared Memory__

- Multiple cores, same memory
- Fast communication

__Distributed Memory__

- Multiple machines
- Network communication

# Parallel R

- there are various packages for High Performance Computing in R, <br>
e.g. CRAN view task at https://cran.r-project.org/web/views/HighPerformanceComputing.html

- some of these packages are experimental and only available as github packages



# Base R: `parallel`

```{r}
library(parallel)

detectCores()

```
Main function: switch `lapply` to `mcapply` (for multi-core lapply)

Doesn't work on Windows

```{r}
library(lme4)

f <- function(i) { # i contains core from 1 to k
  lmer(Petal.Width ~ . - Species + (1 | Species), data = iris)
}
```

# Parallelization

Parallel setup and pullback of results are costly overheads

```{r}
system.time(save1 <- lapply(1:120, f)) # single core
system.time(save2 <- mclapply(1:120, f, mc.cores = 2)) # two cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 3)) # three cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 4)) # four cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 5)) # five cores
```

# Diminishing returns

```{r}
system.time(save2 <- mclapply(1:120, f, mc.cores = 6)) # six cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 7)) # seven cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 8)) # eight cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 9)) # nine cores
system.time(save2 <- mclapply(1:120, f, mc.cores = 10)) # ten cores
```


# Cluster Parallelism 

Start separate sockets - also works on Windows

```{r}
cl <- makeCluster(4)
clusterExport(cl, "f")
clusterEvalQ(cl, {
  library(lme4)
})
```

# 

```{r}
system.time({
res <- parLapply(cl, 1:12000, f)
})

stopCluster(cl)
```



# `foreach` + `doParallel`

```{r}
library(foreach)
library(doParallel)

cl <- makeCluster(4)
registerDoParallel(cl)


res <- foreach(i = 1:120, .combine = c, .packages = "lme4") %dopar% {
  f()
}

stopCluster(cl)
```

# `future` Framework

```{r}
library(future)
library(future.apply)

plan(multisession, workers = 4)

system.time(
res <- future_lapply(1:120, f)
)
```

# Performance Considerations

- Overhead matters
- Task granularity
- Memory duplication
- Load balancing

